Developing a Bubble Chamber Particle Discriminator Using Semi-Supervised Learning

### 2018-11-21 documentation/paper.tex
More comments

\begin{enumerate}
    \item A detailed model of the physical environment is often used to produce an accurate discriminator. This takes a long time, and will have to be updated whenever the physical variables of the experimental apparatus change.

@@

\begin{enumerate}
    \item Determine the most effective machine learning architecture to use in development of a discriminator. Furthermore, develop several neural network architectures and apply them to different input/preprocessing formats (such as audio Fourier transforms, raw waveforms, and images), in order to find which are the most effective.

    \item Determine whether these techniques can be extended, applying two original algorithms based on semi-supervised learning, to develop an effective discriminator function based on incomplete and inaccurate initial information.

@@

    \begin{enumerate}
        \item Neutron calibration sets, which are known to contain approximately 90\% neutrons, and
        \item Previously blinded WIMP search runs and diagnostic background runs, which contain approximately 99\% alpha particles.
    \end{enumerate}
\end{enumerate}

The architectures described above may have wider applications to several similar cases, such as those that require a discriminator which is difficult to construct using conventional techniques. The following are possible reasons for this limitation:

@@

The target was superheated by reducing its pressure until it entered a metastable state in which the vapor phase became energetically accessible. In the presence of a sufficiently dense energy deposition, a bubble would be nucleated and would grow to visible size. Four cameras stereoscopically monitored the chamber for these bubbles, issuing a recompression trigger upon their appearance. At this time, acoustic traces were recorded that captured the sound of bubble formation, and a Dytran 2006V1 fast pressure transducer \cite{amole_thesis} monitored the rise. The rising pressure caused the vapor bubble to condense as the fluid reverted to a non-superheated state. Following this, the hydraulic system prepared to return the detector to the sensitive superheated state by once again expanding to a lower pressure.

Alpha decays along the U/Th chains present the dominant remaining background which must be distinguished from nuclear recoils acoustically \cite{isotopes}.

@@

The Acoustic Parameter method (furthermore AP) is the technique that was used for event discrimination in the original PICO-60 analysis. A function was defined that discriminates between alpha particles and nuclear recoils based on audio data. It was derived based on a physical model of the bubble chamber, and was tuned using neutron calibration sources to produce events in the PICO-60 apparatus.

@@

Discriminators can use a variety of different data formats to make predictions. These are collected from the two essential sensors present in the PICO-60 apparatus: piezoelectric microphones (piezos) and cameras.

@@

The lowest-level data derived from the piezos is the raw waveform $\omega$. This consists of a series of 250,000 samples, collected at the rate of one per microsecond. Each sample is represented as a 16-bit integer. Only the section of the audio between sample 90,000 (inclusive) and sample 190,000 (exclusive) was used during most experiments, because there is only background noise prior to this window, and a clipped signal produced by hydraulics repressurizing the vessel after. These sections contain no information and can be undesirably fit on by neural networks. (For context, approximately half of the samples are recorded prior to the camera trigger.)

@@

The raw waveform $\omega$ was converted into the frequency domain by means of a one-dimensional Discrete Fourier Transform (DFT) for real input. Since DFTs produce sequences of complex numbers, the magnitude of each element was computed. The direct output from such a DFT is the full-resolution Fourier transform $\beta _{50,001}$, which consists of a sequence of numbers half the length of the raw waveform $\omega$.

Beyond this, the banded Fourier transform $\beta_{8}$ (which is the input to the AP function and the original machine learning analysis) was computed by integrating the resonant energy over all frequencies within each of a set of frequency bands ranging from 1 kHz to 300 kHz \cite{pico}. This can be thought of as a method of signal downscaling, compressing the information into a smaller number of data points.

@@

Bubble events are detected using the four cameras within the PICO-60 apparatus, with an image-based trigger. When this happens, each camera records a sequence of 41 images before and after formation. These raw images contain a large amount of extraneous information; they encompass the entire vessel. To reduce the input information, the image window sequence $\iota$ includes 50\texttimes50 cropped windows around the position of the bubble. The 41 frames, many of which contain either no bubble or a bubble in later stages of formation, are reduced to 10 immediately around the formation of the bubble. Five frames from before the recording trigger are included in $\iota$, because the bubble does not cause a trigger until it is already of a significant size.

@@

The 3-dimensional position $\chi$ of the bubble within the vessel is calculated using triangulation, based on the known positions and angles of the cameras and the position of the bubble within the field of view of each camera. This is used in the banded frequency position correction function $PosCor(\beta _{8}, \chi)$, which corrects $\beta _{8}$ for variations in amplitude which depend on the position of the bubble.

@@

The initial data set $D$ to which these cuts are applied consists of all events recorded in the PICO-60 detector between 2016-09-02 and 2017-06-20.

@@

PICO-60 events which include multiple bubbles are \textit{always} neutron events; alpha particles are stopped electromagnetically, and WIMP candidates have a negligible probability of multiply scattering because of their extremely low predicted cross section. Thus, multi-bubble events can be safely removed. The bubble multiplicity cut {\it MultiCut}$(D)$ consists of the following restrictions:

@@

Please note that events appearing to be in the middle of the vessel in Figure \ref{wall_cuts_visualization} are in actually at the top and bottom of the detector, as only the X and Y axes are displayed.

@@

Performance of machine learning systems was evaluated using two metrics: classification accuracy and class-wise standard deviation.

@@

The first step is to calculate the standard deviation $S$ of the union of $N$ and $A$. This gives an indication of the scale of the overall distribution. When $N$ and $A$ are divided by $S$, they are normalized so that the standard deviation of their union is equal to 1. While the neural network’s outputs are bounded in the range of 0 to 1 with a sigmoid activation, AP has a significantly wider range. Normalization of the union prevents this from creating a bias where AP would produce a higher standard deviation with a similarly proportioned error.

@@

The following hyperparameters were tested during a grid search:
\begin{itemize}
    \item L2 Regularization $\lambda \in \{0.0003, 0.001, 0.003\}$
    \item Dense Layer Dropout $\in \{0, 0.25, 0.5\}$
    \item Convolutional Filters $\in \{24, 48\}$
    \item Convolutional Kernel Size $\in \{3, 5\}$
    \item Number of Convolutional Layers $\in \{17, 32\}$
\end{itemize}

@@

\subsection{Multi-Layer Perceptron for Fourier Transform Analysis}

@@

When $\beta_{8}$ was used as input, position corrections were optionally applied to the input data, as they were the input to Acoustic Parameter. The resulting configurations are furthermore referred to as $FourierMLP(\beta_{N})$ and $FourierMLP(PosCor(\beta_{N}, \chi))$. While a variety of network architectures were tested, most had a small number of layers (on the order of three), applied dropout and L2 regularization, and used batch normalization once again.

@@

It is very clear that higher resolution data does not improve accuracy. Regularization, both dropout and L2, appears to have relatively little effect.

>>>>

\begin{enumerate}
    \item A detailed model of the physical environment is often used to produce an accurate discriminator. This takes a long time to develop and compute, and will have to be updated whenever the physical variables of the experimental apparatus change.

@@

\begin{enumerate}
    \item Determine the most effective machine learning architecture to use in development of a discriminator. Furthermore, develop several neural network architectures and apply them to different input/preprocessing formats (such as audio Fourier transforms, raw waveforms, and images), in order to find which are the most effective.
    \item Determine whether these techniques can be extended, applying two original algorithms based on semi-supervised learning, to develop an effective discriminator function based on incomplete and inaccurate initial information.
\end{enumerate}

@@

\begin{enumerate}
    \item Neutron calibration sets, which are known to contain approximately 90\% neutrons, and
    \item Previously blinded WIMP search runs and diagnostic background runs, which contain approximately 99\% alpha particles.
\end{enumerate}

The architectures described in this paper may have wider applications to several similar cases, such as those that require a discriminator which is difficult to construct using conventional techniques. The following are possible reasons for this limitation:

@@

The target was superheated by reducing its pressure until it entered a metastable state in which the vapor phase became energetically accessible. In the presence of a sufficiently dense energy deposition, a bubble would be nucleated and would grow to visible size. Four cameras stereoscopically monitored the chamber for these bubbles, issuing a recompression trigger upon their appearance. At this time, acoustic traces were recorded that captured the sound of bubble formation, and a Dytran 2006V1 fast pressure transducer \cite{amole_thesis} monitored the pressure rise. The rising pressure caused the vapor bubble to condense as the fluid reverted to a non-superheated state. Following this, the hydraulic system prepared to return the detector to the sensitive superheated state by once again expanding to a lower pressure.

Superheated fluid detectors have an intrinsic insensitivity to the relatively diffuse energy depositions from electron recoils \cite{pico}, and neutron backgrounds are mitigated by shielding the detector volume in a 20 t water tank. Alpha decays along the U/Th chains present the dominant remaining background which must be distinguished from nuclear recoils acoustically \cite{isotopes}.

@@

An Acoustic Parameter method (furthermore AP) is the technique that was used for event discrimination in the original PICO-60 analysis. A function was defined that discriminates between alpha particles and nuclear recoils based on audio data. It was derived based on a physical model of the bubble chamber, and was tuned using neutron calibration sources to produce events in the PICO-60 apparatus.

@@

Discriminators can use a variety of different data formats to make predictions. These are collected from the two essential sensors present in the PICO-60 apparatus: two piezoelectric microphones (piezos) and cameras.

@@

The lowest-level data derived from the two piezos in the PICO-60 apparatus is the raw waveform $\omega$. This consists of a series of 250,000 samples, collected at the rate of one per microsecond. Each sample is represented as a 16-bit integer. Only the section of the audio between sample 90,000 (inclusive) and sample 190,000 (exclusive) was used during most experiments, because there is only background noise prior to this window, and a clipped signal produced by hydraulics repressurizing the vessel after. These sections contain no information and can be undesirably fitted by neural networks. (For context, approximately half of the samples are recorded prior to the camera trigger.)

@@

The raw waveform $\omega$ was converted into the frequency domain by means of a one-dimensional Discrete Fourier Transform (DFT) for real input. Since DFTs produce sequences of complex numbers, the magnitude of each element was computed. The direct output from such a DFT is the full-resolution Fourier transform $\beta _{50,001}$, which consists of a sequence of 50,001 numbers (half the length of the raw waveform $\omega$).

Beyond this, the banded Fourier transform $\beta_{8}$ (which is the input to the AP function and the original machine learning analysis) was computed by integrating the resonant energy over all frequencies within each of a set of eight frequency bands ranging from 1 kHz to 300 kHz \cite{pico}. This can be thought of as a method of signal downscaling, compressing the information into a smaller number of data points.

Both $\beta _{50,001}$ and $\beta_{8}$ were used as inputs to machine learning models applied in this study.

@@

Bubble events are detected using the four cameras within the PICO-60 apparatus, with an image-based trigger. When this happens, each camera records a sequence of 41 images before and after formation. These raw images contain a large amount of extraneous information; they encompass the entire vessel. To reduce the input information, the image window sequence $\iota$ includes 50\texttimes50 cropped windows around the position of the bubble. The 41 frames, many of which contain either no bubble or a bubble in later stages of formation, are reduced to 10 frames immediately around the formation of the bubble. Two to three frames from before the recording trigger are included in $\iota$, because the bubble does not cause a trigger until it is already of a significant size.

@@

The 3-dimensional position $\chi$ of the bubble within the vessel is calculated using triangulation, based on the known positions and angles of the cameras and the position of the bubble within the field of view of each camera.

This is used in the banded frequency position correction function $PosCor(\beta _{8}, \chi)$, which corrects $\beta _{8}$ for variations in amplitude which depend on the position of the bubble. $PosCor(\beta _{8}, \chi)$ is used for training several machine learning models throughout this study. Details of its use are in Section \ref{mlp}.

@@

The initial data set $D$ to which these cuts are applied consists of all events recorded in the PICO-60 detector, at the 3.3keV threshold, between 2016-09-02 and 2017-06-20.

@@

PICO-60 events which include multiple bubbles are \textit{always} neutron events; alpha particles are stopped electromagnetically, and WIMP candidates have a negligible probability of multiple scattering because of their extremely low predicted cross section. Thus, multi-bubble events can be safely removed. The bubble multiplicity cut {\it MultiCut}$(D)$ consists of the following restrictions:

@@

Some events eliminated by fiducial cuts appear to be in the middle of the vessel in Figure \ref{wall_cuts_visualization}. These events are at the top or bottom of the detector (along the Z axis, which is not displayed in the 2D plot).

The data set, after all of these cuts are applied, consists of 624 events. During each training run, 128 randomly selected events are set aside for validation, leaving a training set of 496 events.

@@

Performance of the machine learning systems was evaluated using two metrics: classification accuracy and class-wise standard deviation.

@@

The first step is to calculate the standard deviation $S$ of the union of $N$ and $A$. This gives an indication of the scale of the overall distribution. When $N$ and $A$ are divided by $S$, they are normalized so that the standard deviation of their union is equal to 1. While the neural network’s outputs are bounded in the range of 0 to 1 with a sigmoid activation on the last layer, AP has a significantly wider range. Normalization of the union prevents this from creating a bias where AP would produce a higher standard deviation with a similarly proportioned error.

@@

The following hyperparameters were tested during a grid search:
\begin{itemize}
    \item L2 Regularization $\lambda \in \{0.0003, 0.001, 0.003\}$
    \item Dense Layer Dropout $\in \{0, 0.25, 0.5\}$
    \item Convolutional Filters $\in \{24, 48\}$
    \item Convolutional Kernel Size $\in \{3, 5\}$
    \item Number of Convolutional Layers $\in \{17, 32\}$
    \item Number of Dense Layers $= 3$
\end{itemize}

@@

\subsection{Multi-Layer Perceptron for Fourier Transform Analysis} \label{mlp}

@@

When $\beta_{8}$ was used as input, position corrections were optionally applied to the input data, as they were the input to Acoustic Parameter. The resulting configurations are furthermore referred to as $FourierMLP(\beta_{N})$ and $FourierMLP(PosCor(\beta_{N}, \chi))$. While a variety of network architectures were tested, most had a small number of dense layers (on the order of three), applied dropout and L2 regularization, and used batch normalization once again.

@@

It is very clear that higher resolution data does not improve accuracy in a relatively shallow network. Regularization, both dropout and L2, appears to have relatively little effect.
